A Switching Planner for Combined Task and Observation Planning

### 2011-02-09 code/subarchitectures/planner.sa/trunk/src/latex/AAAI11/switching.tex
Typos and minor edits in a number of places. Switching continual planner section clarified and shortened.

We now describe our {\em switching} planning system that operates
according to the continual planning paradigm, given a description of
the current problem and domain in DTPDDL. The system {\em
switches}, in the sense that planning proceeds in interleaved
sessions, in which the base planner is either {\em sequential}/linear
or {\em decision-theoretic}.
%%
During a sequential session, a rewarding {\em trace} of a possible
execution is computed using a cost-optimising satisficing planner
which trades action costs, goal rewards, and determinacy.
%%
Formatted as a linear plan, the trace specifies a sequence of actions
that achieves some reachable goals following a deterministic
approximation of the problem at hand.
%%
Structurally, a trace is a sequence of elements that are either: (i) ground
actions from the DTPDDL description of the world, or (ii) atomic {\em
assumptions}, modelled as deterministic actions, made about the truth
value of facts that can only be determined at runtime (e.g., that a
box of cornflakes is located on the corner bench in the kitchen).
%%
The system always begins in a sequential session, so that plan
execution proceeds by applying DTPDDL actions from the trace, in
sequence, until the outcome of executing the next scheduled action is
too uncertain according to a threshold parameter~(here, 95\%). In that
eventuality a DT session commences, which tailors sensory processing
to determine whether the assumptions made in the trace hold, or which
otherwise acts to achieve the goals.

Because online DT planning in large problems is impractical (we seek
response times in seconds), a DT session plans for an abstract process
determined by the current trace and underlying belief-state. That
abstract problem is characterised by a limited number of propositions,
chosen due to their {\em relevance} to the action whose scheduled
execution triggered the switch to the DT session.
%%the subproblem the DT session is asked tosolve.
% MOG: Describe this in detail in the appropriate subsection
%  This
% abstraction is constructed by first excluding all propositions (and
% related actions) that are not true, or {\em assumed } true, of states
% in the trace, then adding them back, using as an heuristic the entropy
% of the trace assumptions conditional on a candidate
% proposition. Propositions are added, one at a time, until the number
% of states in the initial belief-state reaches a given
% threshold.\footnote{Our experiments are performed using a number of
% different threshold parameters.} 
To that abstract model we also add {\em disconfirm} and {\em confirm}
actions that the DT session can schedule in order to judge an atomic
assumption in the trace. Those actions yield a relatively small reward
if the corresponding judgement is true (or small penalty
otherwise). If a judgement action is scheduled for execution the DT
session is terminated, and control is returned to a new sequential
session.

Whatever the session type, our continual planner maintains a factored
representation of successive belief-states by performing belief
revision. As an internal representation of the $(\pp{:init})$
declaration, we keep a tree-shaped Bayesian network which gets updated
whenever an action is performed, or an observation received. That
belief-state representation is used: (1) as the source of candidate
determinisations for sequential planning, (2) in determining when to
switch to a DT session, and (3) as a mechanism to guide construction
of an abstract process for DT sessions.

@@

\noindent Each assumption corresponds to one distinct {\em
relaxed} visitation of the root term. Here, a conjunctive term is
visited iff its atomic subterms are visited, and zero or one of its
immediate probabilistic subterms are visited. For a sequential
session, the starting state corresponds to an assumption (usually
unique) with probability $1$.\footnote{To simplify the discussion, we
suppose that the root term of $(\pp{:init})$ lists all atoms that are
necessarily true.} In our example, that starting state is:

@@

Encapsulating state-assumptions, we annotate the problem posed during
a sequential session with an \emph{assumptive action} $\assumptiveS{i}$ for
each element $\prob_i (T_i)$, of each probabilistic term from
$(\pp{:init})$. Here, $\assumptiveS{i}$ can be executed if no
$\assumptiveS{j}$, $j \neq i$, has been executed from the same
probabilistic term, and, either
$(\pp{probabilistic}~..\prob_i~(T_i)..)$ is in the root conjunct, or
it occurs in $T_k$ for some executed $\assumptiveS{k}$.
%%
We also add constraints that forbid scheduling of
assumptions about facts after actions with preconditions or effects
that mention those facts. For example, the robot cannot assume it is
plugged into a power source immediately after it unplugs itself.
%%
Executing $\assumptiveS{i}$ in a state $\state$ effects a transition
to a successor state $\state^{T_i}$ with probability $\prob_i$, and
$\state^\bot$ with probability $1 - \prob_i$. Here, $\state^{T_i}$ is
the union of $\state$ with atomic terms from $T_i$. State
$\state^\bot$ is an added sink.

We now describe the optimisation criteria used during sequential
sessions. Where $\prob_i$ is the probability that the $i^{th}$
sequenced action, $\action_i$, from a trace of state-action pairs
$\langle \state_0, \action_0,\state_1, \action_1,.., \state_N \rangle$
does not transition to $\state^\bot$, we have that the optimal trace
has value:

@@

\noindent  Here,  $\reward(\state_i, \action_i)$ is the instantaneous
reward ---i.e., can be negative for action costs, or positive for goal
achievement--- received for executing action $\action_i$ in state
$\state_i$. Finally, it is worth clarifying that in practise we do not
artificially limit the length of traces that the sequential planner
can consider. Moreover, for problems we consider there is always a
finite optimal trace.

@@

In order to guarantee fast DT sessions, planning occurs for a small
abstract problem. In our setting, where the underlying decision
process has a deterministic action model, a significant cause of
problem largeness is due to the number of states that occur with
non-zero probability in the underlying belief-state. In posing a small
abstract process for a DT session, we construct an abstract
belief-state that is chiefly characterised by facts related to the
current trace. In that abstraction, instances of operator or sense
declarations whose specification mentions an excluded fact are not
available.
%%
In detail, we first construct an $(\pp{:init})$ declaration which
contains the root term, and relevant assumptions from the trace --
e.g., (Fig.\ref{fig:abstraction-b}) gives an example fragment, where
diamonds are probabilistic terms, and circles are atomic and/or
conjunctive.
%%
For each excluded atom, we compute the {\em entropy} of the active
assumptions, {\em conditional} on that atom. Intuitively, lower
entropy indicates an atom gives better information about
assumptions. Facts are iteratively added to the belief-state in
increasing order according to that measure until the belief space size
reaches a predefined limit (Fig.\ref{fig:abstraction-c}).

>>>>

We now describe our {\em switching} planning system that operates
according to the continual planning paradigm given a description of
the current problem and domain in DTPDDL. The system {\em
switches} in the sense that planning proceeds in interleaved
sessions in which the base planner is either {\em sequential}/linear
or {\em decision-theoretic}.
%%
The system begins in a sequential session during which it runs a
cost-based satisficing planner which trades off action costs, goal
rewards and determinacy. This planner generates a reward-giving
sequence of ground actions from the DTPDDL domain which may also
include {\em assumptions}: actions that assume the truth of some fact
that is unknown at plan time (e.g. that a box of cornflakes is located
on the corner bench in the kitchen). We call this sequence a {\em
trace}. Execution of the trace proceeds by applying the actions until
the probability of the next action being applicable falls below some
threshold (here 95\%), at which point the planner switches to a DT session.

Because online DT planning is impractical for the size of problem we
are interested in, when a DT session begins we use the trace from the
preceeding sequential session to restrict the state space. The state space is characterised by a limited number of propositions,
chosen due to their {\em relevance} to the action whose scheduled
execution triggered the switch to the DT session.
%The state
%space is constructed by first removing all propositions (and related
%actions) that are not true or assumed true of states in the trace,
%then adding them back as long as the number of states does not exceed
%a size threshold\footnote{Our experiments are performed using a number
%of different threshold parameters.}  Propositions to add back are
%chosen using as a heuristic the entropy of the trace assumptions
%conditional on the candidate proposition. 
To allow the DT planner to
determine the truth of the assumptions in the trace, we add {\em
disconfirm} and {\em confirm} actions to the problem for each of
them. These actions yield a relatively small reward/penalty if the
corresponding judgement is true/false. If a judgement action is
scheduled for execution the DT session is terminated, and control is
returned to a sequential session.

%% We now describe our {\em switching} planning system that operates
%% according to the continual planning paradigm, given a description of
%% the current problem and domain in DTPDDL. The system {\em
%% switches}, in the sense that planning proceeds in interleaved
%% sessions, in which the base planner is either {\em sequential}/linear
%% or {\em decision-theoretic}.
%% %%
%% During a sequential session, a rewarding {\em trace} of a possible
%% execution is computed using a cost-optimising satisficing planner
%% which trades action costs, goal rewards, and determinacy.
%% %%
%% Formatted as a linear plan, the trace specifies a sequence of actions
%% that achieves some reachable goals following a deterministic
%% approximation of the problem at hand.
%% %%
%% Structurally, a trace is a sequence of elements that are either: (i) ground
%% actions from the DTPDDL description of the world, or (ii) atomic {\em
%% assumptions}, modelled as deterministic actions, made about the truth
%% value of facts that can only be determined at runtime (e.g., that a
%% box of cornflakes is located on the corner bench in the kitchen).
%% %%
%% The system always begins in a sequential session, so that plan
%% execution proceeds by applying DTPDDL actions from the trace, in
%% sequence, until the outcome of executing the next scheduled action is
%% too uncertain according to a threshold parameter~(here, 95\%). In that
%% eventuality a DT session commences, which tailors sensory processing
%% to determine whether the assumptions made in the trace hold, or which
%% otherwise acts to achieve the goals.

%% Because online DT planning in large problems is impractical (we seek
%% response times in seconds), a DT session plans for an abstract process
%% determined by the current trace and underlying belief-state. That
%% abstract problem is characterised by a limited number of propositions,
%% chosen due to their {\em relevance} to the action whose scheduled
%% execution triggered the switch to the DT session.
%% %%the subproblem the DT session is asked tosolve.
%% % MOG: Describe this in detail in the appropriate subsection
%% %  This
%% % abstraction is constructed by first excluding all propositions (and
%% % related actions) that are not true, or {\em assumed } true, of states
%% % in the trace, then adding them back, using as an heuristic the entropy
%% % of the trace assumptions conditional on a candidate
%% % proposition. Propositions are added, one at a time, until the number
%% % of states in the initial belief-state reaches a given
%% % threshold.\footnote{Our experiments are performed using a number of
%% % different threshold parameters.} 
%% To that abstract model we also add {\em disconfirm} and {\em confirm}
%% actions that the DT session can schedule in order to judge an atomic
%% assumption in the trace. Those actions yield a relatively small reward
%% if the corresponding judgement is true (or small penalty
%% otherwise). If a judgement action is scheduled for execution the DT
%% session is terminated, and control is returned to a new sequential
%% session.

Whatever the session type, our continual planner maintains a factored
representation of successive belief-states.
% by performing belief revision. 
As an internal representation of the $(\pp{:init})$
declaration, we keep a tree-shaped Bayesian network which gets updated
whenever an action is performed, or an observation received. That
belief-state representation is used: (1) as the source of candidate
determinisations for sequential planning, (2) in determining when to
switch to a DT session, and (3) as a mechanism to guide construction
of an abstract process for DT sessions.

@@

\noindent Each assumption corresponds to one distinct {\em
relaxed} visitation of the root term. Here, a conjunctive term is
visited iff its atomic subterms are visited, and zero or one of its
immediate probabilistic subterms are visited. For a sequential
session, the starting state corresponds to an assumption (usually
unique) with probability $1$.\footnote{Without a loss of generality,
we assume conjunctive terms list atoms that are true with
probability 1 according to a visitation, and that actions do not
have negative preconditions.}  In our example, that starting state is:

@@

To represent state assumptions we augment the problem posed during
a sequential session with an \emph{assumptive action} $\assumptiveS{i}$ for
each element $\prob_i (T_i)$, of each probabilistic term from
$(\pp{:init})$. Here, $\assumptiveS{i}$ can be executed if no
$\assumptiveS{j}$, $j \neq i$, has been executed from the same
probabilistic term, and, either
$(\pp{probabilistic}~..\prob_i~(T_i)..)$ is in the root conjunct, or
it occurs in $T_k$ for some executed $\assumptiveS{k}$.
%%
We also add constraints that forbid scheduling of
assumptions about facts after actions with preconditions or effects
that mention those facts. For example, the robot cannot assume it is
plugged into a power source immediately after it unplugs itself.
%%
Executing $\assumptiveS{i}$ in a state $\state$ effects a transition
to a successor state $\state^{T_i}$ with probability $\prob_i$, and
$\state^\bot$ with probability $1 - \prob_i$. Here, $\state^{T_i}$ is
the union of $\state$ with atomic terms from $T_i$. State
$\state^\bot$ is an added sink.

We now describe the optimisation criteria used during sequential
sessions. Let $\prob_i$ be the probability that the $i^{th}$ sequenced
action, $\action_i$, from a trace of state-action pairs
$\langle \state_0, \action_0,\state_1, \action_1,.., \state_N \rangle$
does not transition to $\state^\bot$, and
$\reward(\state_i, \action_i)$ be the instantaneous reward received
for executing action $\action_i$ in state $\state_i$. The optimal
trace has value:

%We now describe the optimisation criteria used during sequential
%sessions. Where $\prob_i$ is the probability that the $i^{th}$
%sequenced action, $\action_i$, from a trace of state-action pairs
%$\langle \state_0, \action_0,\state_1, \action_1,.., \state_N \rangle$
%does not transition to $\state^\bot$, we have that the optimal trace
%has value:

@@

%\noindent  Here,  $\reward(\state_i, \action_i)$ is the instantaneous
%reward ---i.e., can be negative for action costs, or positive for goal
%achievement--- received for executing action $\action_i$ in state
%$\state_i$. Finally, it is worth clarifying that in practise we do not
%artificially limit the length of traces that the sequential planner
%can consider. Moreover, for problems we consider there is always a
%finite optimal trace.

@@

In order to guarantee fast DT sessions, planning occurs for a small
abstract problem. In our setting, where the underlying decision
process has a deterministic action model, a significant cause of
problem largeness is due to the number of states that occur with
non-zero probability in the underlying belief-state. In posing a small
abstract process for a DT session, we construct an abstract
belief-state that is chiefly characterised by facts related to the
current trace. In that abstraction, instances of operator or sense
declarations whose specification mentions an excluded fact are not
available.
%%
In detail, we first construct an $(\pp{:init})$ declaration which
contains the root term, and relevant assumptions from the trace---e.g., (Fig.\ref{fig:abstraction-b}) gives an example fragment, where
diamonds are probabilistic terms, and circles are atomic and/or
conjunctive.
%%
For each excluded atom, we compute the {\em entropy} of the {\em
relevant}
assumptions, {\em conditional} on that atom. Intuitively, lower
entropy indicates an atom gives better information about
assumptions. Facts are iteratively added to the belief-state in
increasing order according to that measure until the belief space size
reaches a predefined limit (Fig.\ref{fig:abstraction-c}).
